---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# treenetproc

<!-- badges: start -->
<!-- badges: end -->

The package `treenetproc` cleans raw dendrometer data. The cleaning process 
aligns measurements at regular time intervals and removes jumps and outliers.
All changes are flagged and can be visually examined in the graphical outputs.

## Installation

You can install the latest stable version of the package from 
[GitHub](https://github.com/) with:

``` {r install packge, echo = TRUE, results = "hide", message=FALSE, warning=FALSE}
# install.packages("devtools")
library(devtools)
devtools::install_github("treenet/treenetproc")
```
<br>


## Example

Below you can see a basic example of a processing result. 

```{r Fig_1, fig.width = 6, fig.height = 2, fig.retina = 2, echo = FALSE}
library(treenetproc)
data("dendro_data_L1")
data("dendro_data_L2")
example_L1 <- dendro_data_L1 %>% 
  dplyr::filter(series == "site-1_dendro-1")
example_L2 <- dendro_data_L2 %>% 
  dplyr::filter(series == "site-1_dendro-1")
layout(matrix(c(1, 2), nrow = 2), heights = c(1, 1),
       widths = 1)
par(mar = c(0, 2.1, 0, 1))
plot(data = example_L1, value ~ ts, type = "l", xaxt = "n", 
     yaxt = "n", ylab = "", las = 1, col = "grey70")
title(ylab = "raw", mgp = c(0.8, 1, 0))
par(mar = c(0, 2.1, 0, 1))
plot(data = example_L2, value ~ ts, type = "n", xaxt = "n", 
     yaxt = "n", ylab = "", xlab = "")
title(ylab = "processed", mgp = c(0.8, 1, 0))
lines(data = example_L2, value ~ ts, col = "#08519c")
```
<br>

For more information on the functionality of the package please see the 
package vignettes:

```{r install vignettes, results="hide", message=FALSE, warning=FALSE}
browseVignettes("treenetproc")
```
If you want to access the vignettes you need to build them when installing
the package:

``` {r install package with vignettes, echo = TRUE, results = "hide", message=FALSE, warning=FALSE}
# install.packages("devtools")
library(devtools)
devtools::install_github("treenet/treenetproc", build_vignettes = TRUE)
```
<br>


## Citation
To cite `treenetproc` in a publication use:

```{r citation, echo = TRUE, message=FALSE, warning=FALSE}
citation("treenetproc")
```
