---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# treenetproc

<!-- badges: start -->
<!-- badges: end -->

The package `treenetproc` cleans raw dendrometer data. The cleaning process 
aligns measurements at regular time intervals, removes jumps and outliers,
flags all changes and plots them as well.

## Installation

You can install the development version of the package from 
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("treenet/treenetproc")
```

## Example

Below you can see a basic example of a processing result. 

```{r Figure 1, fig.width = 6, fig.height = 2, fig.retina = 2, echo = FALSE}
library(treenetproc)
data("dendro_data_L1")
data("dendro_data_L2")
example_L1 <- dendro_data_L1 %>% 
  dplyr::filter(series == "site-1_dendro-1")
example_L2 <- dendro_data_L2 %>% 
  dplyr::filter(series == "site-1_dendro-1")
layout(matrix(c(1, 2), nrow = 2), heights = c(1, 1),
       widths = 1)
par(mar = c(0, 2.1, 0, 1))
plot(data = example_L1, value ~ ts, type = "l", xaxt = "n", 
     yaxt = "n", ylab = "", las = 1, col = "grey70")
title(ylab = "raw", mgp = c(0.8, 1, 0))
par(mar = c(0, 2.1, 0, 1))
plot(data = example_L2, value ~ ts, type = "n", xaxt = "n", 
     yaxt = "n", ylab = "", xlab = "")
title(ylab = "processed", mgp = c(0.8, 1, 0))
lines(data = example_L2, value ~ ts, col = "#08519c")
```

For more information on the functionality of the package please see the 
package vignettes:

```{r browse vignettes, results="hide", message=FALSE, warning=FALSE}
browseVignettes("treenetproc")
```
