---
title: "Introduction to treenetproc"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    theme: paper
vignette: >
  %\VignetteIndexEntry{Introduction to treenetproc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.asp = 0.618,
  out.widht = "80%"
)
# This is necessary to direct knitr to find the 
# 'data' directory, thanks to 
# https://stackoverflow.com/a/24585750/1036500
knitr::opts_knit$set(root.dir = normalizePath('../'))
```

<br>
```{r Figure 1, fig.asp = 0.35, fig.retina = 2, echo = FALSE}
library(treenetproc)
data("dendro_data_L1")
data("dendro_data_L2")
example_L1 <- dendro_data_L1 %>% 
  dplyr::filter(series == "site-1_dendro-1") %>% 
  dplyr::filter(ts <= "2013-06-01")
example_L2 <- dendro_data_L2 %>% 
  dplyr::filter(series == "site-1_dendro-1") %>% 
  dplyr::filter(ts <= "2013-06-01")
layout(matrix(c(1, 2), nrow = 2), heights = c(1, 1),
       widths = 1)
par(mar = c(0, 2.1, 0, 1))
plot(data = example_L1, value ~ ts, type = "l", xaxt = "n", 
     yaxt = "n", ylab = "", las = 1, col = "grey70")
title(ylab = "raw", mgp = c(0.8, 1, 0))
par(mar = c(0, 2.1, 0, 1))
plot(data = example_L2, value ~ ts, type = "n", xaxt = "n", 
     yaxt = "n", ylab = "", xlab = "")
title(ylab = "processed", mgp = c(0.8, 1, 0))
lines(data = example_L2, value ~ ts, col = "#08519c")
```
<br><br>

```{r setup, warning = FALSE, message = FALSE}
library(treenetproc)
library(dplyr)
```

The package `treenetproc` cleans dendrometer data by removing outliers and 
jumps in the data. The data is first aligned to regular time intervals 
(level 1: `L1`) and then cleaned (level 2: `L2`).
<br><br>


# Format input data
The raw dendrometer or climate data has to contain a timestamp column named
`ts`. The timestamp can be in any standard date format (default is 
`date_format = "%Y-%m-%d %H:%M:%S"`).

Raw dendrometer or climate data can be provided in long format with 
dendrometer names specified in a column named `series` (Table 1). 
Alternatively the data can be provided in wide format with data series in 
different columns (Table 2).
<br><br>

```{r Table 1 - head of dendro_data_L0, echo = FALSE, results = "asis"}
data("dendro_data_L0")
knitr::kable(dendro_data_L0[1:5, ], caption = "**Table 1**: Raw dendrometer 
             data in `long` format.")
```

<br>

```{r Table 2 - head of dendro_data_L0_wide, echo = FALSE, results = "asis"}
data("dendro_data_L0_wide")
knitr::kable(dendro_data_L0_wide[1:5, ], caption = "**Table 2**: Raw 
             dendrometer and temperature data in `wide` format.")
```
<br>


# Time-alignment (L1)
After converting the data to the correct format (`long` or `wide`), it has 
to be time-aligned to regular time intervals with the function `proc_L1`. 
The resulting time resolution can be specified with the argument `reso` (in
minutes).  
How the time alignment works is documented in 
([Functionality of treenetproc](Functionality-of-treenetproc.html), this 
vignette is under development and not yet available).

```{r prepare data for proc_L1, echo = FALSE}
# reduce size of dendro_data_L0 to make computing of proc_L1 faster
data("dendro_data_L0")
dendro_data_L0 <- dendro_data_L0[1:100, ]
```

```{r examples of proc_L1, results = "hide"}
# Data in long format (cf. Table 1)
proc_L1(data_L0 = dendro_data_L0, reso = 10, input = "long")

# Data in wide format (cf. Table 2)
proc_L1(data_L0 = dendro_data_L0_wide, reso = 10, input = "wide")
```
<br>


# Data cleaning and processing (L2)
Time-aligned dendrometer data can be cleaned with the function 
`proc_dendro_L2`.

```{r prepare for proc_dendro_L2, echo = FALSE}
# reduce size of dendro_data_L1 to make computing of proc_dendro_L2 faster
data("dendro_data_L1")
dendro_data_L1 <- dendro_data_L1 %>% 
  filter(series == "site-1_dendro-1") %>% 
  filter(ts <= "2013-06-01")
```

```{r process to L2, results = "hide", message = FALSE, warning = FALSE}
# Clean time-aligned (L1) dendrometer data
proc_dendro_L2(dendro_data = dendro_data_L1, temp_data = temp_data_L1, 
               plot = FALSE)
```
<br>

## Output data
The function returns a `data.frame` containing the cleaned dendrometer data
(Table 3).

```{r Table 3, echo = FALSE}
data("dendro_data_L2")
knitr::kable(dendro_data_L2[1:5, ], caption = "**Table 3**: Processed 
             dendrometer data.")
```

The output `data.frame` contains the following columns:

* **series**: name of the series
* **ts**: timestamp with format `%Y-%m-%d %H:%M:%S`
* **value**: dendrometer value
* **max**: highest measured dendrometer `value` up to this timestamp
* **twd**: tree water deficit, i.e. the amount of stem shrinkage
  expressed as the difference between `max` and `value` (calculated according
  to Zweifel et al. 2016)
* **mds**: maximum daily shrinkage (see Details section in the help file of
  `proc_dendro_L2()` for more information)
* **gro_yr**: growth since the beginning of the year (calculated according to
  Zweifel et al. 2016)
* **gro_start**: day of year at which growth starts. `gro_start` is defined
  as the day of year at which 5% of total yearly growth is surpassed
* **gro_end**: day of year at which growth stops. `gro_end` is defined as 
  the day of year at which 95% of total yearly growth is reached
* **frost**: indicates periods of probable frost (i.e. periods where the 
  temperature < `lowtemp`)
* **flags**: documentation of all changes that occurred during data cleaning, 
  see details below
* **version**: package version number
<br><br>

The column `flags` documents all changes that occurred during the data 
cleaning. The numbers after the name of the flag specify in which iteration 
of the cleaning process the changes occurred:

* `"out"`: outlier point removed (e.g. `"out1"` for an outlier removed in
  iteration 1 of the cleaning process)
* `"jump"`: jump corrected
* `"fill"`: value was linearly interpolated (length of gaps that are linearly
  interpolated can be specified with the argument `interpol`)
<br><br>

## Output plots
By default all changes are plotted (`plot = TRUE`) and saved to a PDF file
in the current working directory (`plot_export = TRUE`). 

```{r process to L2 first_plot, fig.asp = 0.9, fig.retina = 2, fig.keep = "first", results = "hide", message = FALSE, warning = FALSE}
# Load dataset
data("dendro_data_L1")

# Subset dataset for example
data_L1 <- dendro_data_L1 %>% 
  filter(series == "site-1_dendro-1") %>% 
  filter(ts < "2014-12-31")

# Clean time-aligned (L1) dendrometer data and plot changes
proc_dendro_L2(dendro_data = data_L1, temp_data = temp_data_L1,
               plot_period = "monthly", plot_export = FALSE)
```
<br>

The plot contains...

* the original time-aligned `L1` data in the first panel, 
* the cleaned `L2` data (with `L1` in the background) in the second 
  panel. Interpolated points are circled and frost periods are indicated
  with a horizontal, light-blue line
* the changes between `L1` and `L2` (red) as well as the deleted values
  (pink) in the third panel
* and the tree water deficit (twd) in the last panel
<br>

When plotted monthly (`plot_period = "monthly"`), the changes that occurred 
during the processing are numbered. Alternatively, the plots can be drawn for 
the full period (`plot_period = "full"`) or for each year separately 
(`plot_period = "yearly"`). In these cases, the changes are not numbered.
<br><br>

As a summary for each dendrometer series, an additional plot shows the yearly 
growth curves aligned by the day of year. The first year is shown as a grey
line since growth starts as soon as the measurements start and does not 
necessarily represent sensible values.  
Beneath the plot all input values as well as the package version used for data 
cleaning are printed.


```{r process to L2 last_plot, fig.asp = 1.4, fig.retina = 2, fig.keep = "last", echo = FALSE, results = "hide", message = FALSE}
proc_dendro_L2(dendro_data = data_L1, temp_data = temp_data_L1, 
               plot_period = "monthly", plot_export = FALSE)
```


# Adjust data cleaning

## Rigidity of data cleaning
The rigidity of the data cleaning can be adjusted with the variables
`tol_out` and `tol_jump`. Both variables control the thresholds used for
the classification of outliers and jumps.

An increase of the variable `tol_out` makes sense if values are deleted 
(i.e. classified as outliers) that are not outliers. A  decrease, in contrast,
is appropriate if outliers are not detected as such.
Similarly, an increase of the value of `tol_jump` makes sense if a dendrometer
series is flattened during processing (see example below). A decrease, in
contrast, is appropriate if obvious jumps are not corrected.
<br>


### Example of too rigid data cleaning
```{r shrinking example code, fig.asp = 0.9, fig.show="hide", results="hide", message = FALSE, warning = FALSE}
# Load dataset
data("dendro_data_L1")

# Subset dataset for example
data_L1 <- dendro_data_L1 %>% 
  filter(series == "site-1_dendro-2")

# Clean dendrometer data (tol_jump = 1, tol_out = 10)
proc_dendro_L2(dendro_data = data_L1, temp_data = temp_data_L1, 
               tol_jump = 1, tol_out = 10, plot_export = FALSE)
```

```{r shrinking example plot, fig.asp = 0.5, fig.retina = 2, out.width = "70%", echo=FALSE}
data("dendro_data_L1")
data_L1 <- dendro_data_L1 %>% 
  filter(series == "site-1_dendro-2")

data_shrink_L2 <- proc_dendro_L2(dendro_data = data_L1, 
                                 temp_data = temp_data_L1, 
                                 tol_jump = 1, tol_out = 10, plot = FALSE)

par(mar = c(2.1, 4, 1, 2))
plot(data = data_shrink_L2, value ~ ts, type = "n", ylab = "L2", 
     las = 1)
lines(data = data_L1, value ~ ts, col = "grey70")
lines(data = data_shrink_L2, value ~ ts, col = "#08519c")
```
<br><br>

Such problems can be fixed by increasing the value of `tol_jump`.

<br>
```{r shrinking example higher tol code, fig.asp = 0.9, fig.show="hide", results="hide", message = FALSE, warning = FALSE}
# Clean dendrometer data (use default values of 'tol_jump' and 'tol_out')
proc_dendro_L2(dendro_data = data_L1, temp_data = temp_data_L1, 
               tol_jump = 50, tol_out = 10, plot_export = FALSE)
```

```{r shrinking example higher tol code #2, fig.asp = 0.9, fig.show="hide", echo=FALSE, results = "hide", message = FALSE, warning = FALSE}
data_L2 <- proc_dendro_L2(dendro_data = data_L1, temp_data = temp_data_L1, 
                          tol_jump = 50, tol_out = 10, plot_export = FALSE)
```

```{r shrinking example higher tol plot, fig.asp = 0.5, fig.retina=2, out.width = "70%", echo=FALSE}
par(mar = c(2.1, 4, 1, 2))
plot(data = data_L2, value ~ ts, type = "n", ylab = "L2", 
     las = 1)
lines(data = data_L1, value ~ ts, col = "grey70")
lines(data = data_L2, value ~ ts, col = "#08519c")
```
<br><br>


## Benefit of temperature data
Including temperature data measured on site ensures that frost events are 
not treated as outliers during data cleaning. Frost is assumed as soon as
the temperature drops below the value specified in `lowtemp`. Temperature 
data  can either be provided along with the dendrometer data (series name 
has to  contain `temp`), or it can be provided separately to the argument
`temp_data` in `proc_dendro_L2()`.

In periods of probable frost (temperature < `lowtemp`) the threshold for 
outlier detection is multiplied by `frost_thr`.
<br>


### Example without temperature data
If no tempearture data is provided, a sample temperature dataset is used.
The sample temperature dataset assumes frost in the months December,
January and February; i.e. the temperature is set to 0°C in Dec-Feb and to 
10°C in all other months.

```{r frost no temp data_1, fig.asp = 0.9, fig.show="hide", results="hide", message=FALSE, warning=FALSE}
# Load dataset
data("dendro_data_L1")

# Subset dataset for example
data_L1 <- dendro_data_L1 %>% 
  filter(series == "site-1_dendro-4")

# Clean dendrometer data (without temperature data)
proc_dendro_L2(dendro_data = data_L1, plot_export = FALSE)
```

```{r frost no temp data_2, fig.asp = 0.5, fig.retina = 2, out.width = "70%", echo=FALSE, results="hide", warning=FALSE, message=FALSE}
# Process to L2 without temperature data
data_L2 <- proc_dendro_L2(dendro_data = data_L1, plot = FALSE)

par(mar = c(2.1, 4, 1, 2))
plot(data = data_L2, value ~ ts, type = "n", ylab = "L2", 
     las = 1)
lines(data = data_L1, value ~ ts, col = "grey70")
lines(data = data_L2, value ~ ts, col = "#08519c")
```

The sample temperature dataset does not assume frost after the end of 
February, therefore many values during the frost shrinkage are classified as
outliers and deleted.
<br>


### Example with temperature data
```{r frost with temp data_1, fig.asp = 0.9, fig.show="hide", results="hide", message=FALSE, warning=FALSE}
# Load temperature data
data("temp_data_L0")

# Time-align temperature data
temp_L1 <- proc_L1(data_L0 = temp_data_L0)

# Clean dendrometer data (with temperature data)
proc_dendro_L2(dendro_data = data_L1, temp_data = temp_L1, 
               plot_export = FALSE)
```

```{r frost with temp data_2, fig.asp = 0.5, fig.retina = 2, out.width = "70%", echo=FALSE, results="hide", warning=FALSE, message=FALSE}
# Load dataset
data("temp_data_L1")
data_L2 <- proc_dendro_L2(dendro_data = data_L1, temp_data = temp_data_L1, 
                          plot = FALSE)

par(mar = c(2.1, 4, 1, 2))
plot(data = data_L2, value ~ ts, type = "n", ylab = "L2", 
     las = 1)
lines(data = data_L1, value ~ ts, col = "grey70")
lines(data = data_L2, value ~ ts, col = "#08519c")
# add abline to indicate frost
data_L2_frost <- data_L2 %>% 
  dplyr::rename(value_L2 = value)
plot_frost_period(df = data_L2_frost)
```

Including the temperature dataset reveals that the temperatures were indeed
lower than `lowtemp` during the frost shrinkage. Periods of probable frost 
(i.e. temperature < `lowtemp`) are indicated with a horizontal line in the 
plot.

The increase of the thresholds for outlier detection in periods of probable
frost prevented the classification of the values during the frost shrinkage as
outliers.
<br><br>


# Manual corrections (L3)
If not all errors in data cleaning can be removed by adjusting the default 
values of `tol_jump` and/or `tol_out`, remaining errors can be removed using
the function `corr_dendro_L2()`. There are three possibilities to revert 
wrong changes:

* **remove**: specify the numbers of the changes that should be reverted.
  Remaining changes are renumbered starting at 1.
* **force**: force a jump in the data that was not corrected for by 
  specifying a point in time prior (but close to) where the jump should occur.
* **delete**: delete an entire region of erroneous data by specifying a
  date range. This can also be done for `L1` data with the function 
  `corr_dendro_L1()`.
<br>


## Example of manual corrections
```{r corr_dendro_L2 example_1, fig.asp = 0.9, fig.keep = "first",  fig.retina=2, results="hide", message=FALSE, warning=FALSE}
# Load dataset
data("dendro_data_L1")

# Subset dataset for example
data_L1 <- dendro_data_L1 %>% 
  filter(series == "site-1_dendro-3")

# Clean dendrometer data
data_L2 <- proc_dendro_L2(dendro_data = data_L1, temp_data = temp_data_L1, 
                          plot_period = "monthly", plot_export = FALSE)
```
<br><br>

The following errors remained after data cleaning:

1. Corrections `59-61` are not necessary since the raw data seems to be correct
2. A jump occurring on August 13 is not corrected
3. Not all erroneous data is removed at the beginning (August 1-4)
<br>

Correct the above-described errors:


```{r corr_dendro_L2 example_3, fig.asp = 0.9, fig.keep = "first", fig.retina=2, results="hide", message=FALSE, warning=FALSE}
# Correct described errors
corr_dendro_L2(data_L1 = data_L1, data_L2 = data_L2, 
               remove = 59:61, 
               force = "2013-08-12", 
               delete = c("2013-08-01", "2013-08-04"),
               series = "site-1_dendro-3", plot_export = FALSE)
```

All three errors have been corrected successfully. In addition, the 
remaining corrections were renumbered, starting at 1.

The corrections are also reflected in the returned `data.frame` and all 
changes are added to the column `flags` as:

* `"rem"`: for removed corrections with the argument `remove`
* `"fjump"`: for forced jumps with the argument `force`
* `"del"`: for deleted values with the argument `delete`
<br><br>


# References
[Zweifel R, Haeni M, Buchmann N, Eugster W (2016) Are trees able to grow in 
periods of stem shrinkage? New Phytol. 211:839-49.](https://doi.org/10.1111/nph.13995)


# Citation
To cite `treenetproc` in a publication use:

```{r citation, echo = TRUE, warning=FALSE}
citation("treenetproc")
```

Consider citing all required packages as well. The package 
[`grateful`](https://github.com/Pakillo/grateful) is a helpful resource for 
this.

```{r cite all dependencies, echo = TRUE, results="hide", warning = FALSE, message = FALSE}
# install.packages("devtools")
devtools::install_github("Pakillo/grateful")
```
