---
title: "Introduction to treenetproc"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: false
    theme: paper
vignette: >
  %\VignetteIndexEntry{Introduction to treenetproc}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.asp = 0.618,
  out.widht = "80%"
)
# This is necessary to direct knitr to find the 
# 'data' directory, thanks to 
# https://stackoverflow.com/a/24585750/1036500
knitr::opts_knit$set(root.dir = normalizePath('../'))
```

<br>
```{r Figure 1, fig.asp = 0.35, fig.retina = 2, echo = FALSE}
library(treenetproc)
data("dendro_data_L1")
data("dendro_data_L2")
example_L1 <- dendro_data_L1 %>% 
  dplyr::filter(series == "site-1_dendro-1") %>% 
  dplyr::filter(ts <= "2013-06-01")
example_L2 <- dendro_data_L2 %>% 
  dplyr::filter(series == "site-1_dendro-1") %>% 
  dplyr::filter(ts <= "2013-06-01")
layout(matrix(c(1, 2), nrow = 2), heights = c(1, 1),
       widths = 1)
par(mar = c(0, 2.1, 0, 1))
plot(data = example_L1, value ~ ts, type = "l", xaxt = "n", 
     yaxt = "n", ylab = "", las = 1, col = "grey70")
title(ylab = "raw", mgp = c(0.8, 1, 0))
par(mar = c(0, 2.1, 0, 1))
plot(data = example_L2, value ~ ts, type = "n", xaxt = "n", 
     yaxt = "n", ylab = "", xlab = "")
title(ylab = "processed", mgp = c(0.8, 1, 0))
lines(data = example_L2, value ~ ts, col = "#08519c")
```
<br><br>

```{r setup, warning = FALSE, message = FALSE}
library(treenetproc)
library(dplyr)
```

The package `treenetproc` cleans, processes and visualises highly resolved 
time series of dendrometer data. In two steps, raw dendrometer data is aligned
to regular time intervals (level 1: `L1`) and cleaned (level 2: `L2`). In an
optional third step several commonly used characteristics such as the start and
end of stem growth can be calculated.
This vignette demonstrates the workflow of the package with sample code. 
For more information on the  functionality of the package, the user is referred
to Knüsel et al. (2021).
<br><br>


# Data input
Raw dendrometer and temperature data can be provided as input. Providing
temperature data is optional. However, temperature data increases the quality
of error detection and processing (cf. Step 2: Providing temperature data).

The raw dendrometer or climate data can be provided in long (Table 1) or wide
format (Table 2). Data of multiple sensors can either be specified in a column 
named `series` to separate the sensors (long format, Table 1) or as separate 
columns (wide format, Table 2). Dendrometer data always has to be provided in
`µm`. In addition, a column named `ts` with timestamps in any standard date 
format (default is `date_format = "%Y-%m-%d %H:%M:%S"`) is required.
Dendrometer and temperature data can either be provided in two separate 
data frames or in the same `data.frame`. If temperature data is provided
along with dendrometer data, the name of the temperature data series has to 
contain the string '`temp`' to be identified as temperature data (cf. Table 2).
<br><br>

```{r Table 1 - head of dendro_data_L0, echo = FALSE, results = "asis"}
data("dendro_data_L0")
knitr::kable(dendro_data_L0[1:5, ], caption = "**Table 1**: Raw dendrometer 
             data in `long` format. Dendrometer measurements have to be
             provided in `µm`.")
```

<br>

```{r Table 2 - head of dendro_data_L0_wide, echo = FALSE, results = "asis"}
data("dendro_data_L0_wide")
knitr::kable(dendro_data_L0_wide[1:5, ], caption = "**Table 2**: Raw 
             dendrometer and temperature data in `wide` format. Dendrometer 
             measurements have to be provided in `µm`.")
```
<br><br>


# Step 1: Time-alignment (L1)
After converting the data to the correct format (`long` or `wide`), it has 
to be time-aligned to regular time intervals with the function `proc_L1`. 
The resulting time resolution can be specified with the argument `reso` (in
minutes, i.e. `reso = 10` for a 10-minute resolution). Temperature data can be
time-aligned in the same way as dendrometer data.

In case the data is provided in `wide` format, this has to be specified
with the argument `input = "wide"`.

```{r prepare data for proc_L1, echo = FALSE}
# reduce size of dendro_data_L0 to make computing of proc_L1 faster
data("dendro_data_L0")
dendro_data_L0 <- dendro_data_L0[1:50, ]
data("dendro_data_L0_wide")
dendro_data_L0_wide <- dendro_data_L0_wide[1:50, ]
```

```{r examples of proc_L1, results = "hide"}
# Time-align dendrometer data in long format (cf. Table 1)
proc_L1(data_L0 = dendro_data_L0, reso = 10)

# Time-align dendrometer and temperature data in wide format (cf. Table 2)
proc_L1(data_L0 = dendro_data_L0_wide, reso = 10, input = "wide")
```
<br><br>


# Step 2: Error detection and processing (L2)
Time-aligned dendrometer data can be cleaned and processed with the function 
`proc_dendro_L2`. To increase the quality of error detection,
time-aligned temperature data can optionally be provided to `temp_L1`.

```{r prepare for proc_dendro_L2, echo = FALSE}
# reduce size of dendro_data_L1 to make computing of proc_dendro_L2 faster
data("dendro_data_L1")
dendro_data_L1 <- dendro_data_L1 %>% 
  filter(series == "site-1_dendro-1") %>% 
  filter(ts <= "2013-05-02")
```

```{r process to L2, results = "hide", message = FALSE, warning = FALSE}
# Clean time-aligned (L1) dendrometer data
proc_dendro_L2(dendro_L1 = dendro_data_L1, temp_L1 = temp_data_L1, 
               plot = FALSE)
```
<br>

### Output data
The function returns a `data.frame` containing the cleaned dendrometer data
(Table 3).

```{r Table 3, echo = FALSE}
data("dendro_data_L2")
knitr::kable(dendro_data_L2[1:5, ], caption = "**Table 3**: Cleaned 
             dendrometer data.")
```

The output `data.frame` contains the following columns:

* **series**: name of the dendrometer series
* **ts**: timestamp with format `%Y-%m-%d %H:%M:%S`
* **value**: dendrometer value (µm)
* **max**: highest measured dendrometer `value` up to this timestamp
* **twd**: tree water deficit (µm), i.e. the amount of stem shrinkage
  expressed as the difference between `max` and `value` (calculated according
  to Zweifel et al. 2016)
* **gro_yr**: growth since the beginning of the year (µm; calculated according 
  to Zweifel et al. 2016)
* **frost**: indicates periods of probable frost (i.e. periods where the 
  temperature is below `lowtemp`)
* **flags**: documentation of all changes that occurred during data cleaning, 
  see details below
* **version**: package version number
<br><br>

The column `flags` documents all changes that occurred during the error
detection and processing. The numbers after the name of the flag specify 
in which iteration of the cleaning process the changes occurred:

* `"out"`: outlier point removed (e.g. `"out1"` for an outlier removed in
  iteration 1 of the cleaning process)
* `"jump"`: jump corrected
* `"fill"`: value was linearly interpolated (length of gaps that are linearly
  interpolated can be specified with the argument `interpol`)
<br><br>


## Evaluation of error detection
The visual checking of the results remains an essential step in dendrometer 
data cleaning. By default all changes are plotted (`plot = TRUE`) and saved to
a PDF file in the current working directory (`plot_export = TRUE`). 

```{r process to L2 first_plot, fig.asp = 0.9, fig.retina = 2, fig.keep = "first", results = "hide", message = FALSE, warning = FALSE}
# Load dataset
data("dendro_data_L1")

# Subset dendrometer dataset for example
dendro_L1 <- dendro_data_L1 %>% 
  filter(series == "site-1_dendro-1") %>% 
  filter(ts < "2014-12-31")

# Clean time-aligned (L1) dendrometer data and plot changes
proc_dendro_L2(dendro_L1 = dendro_L1, temp_L1 = temp_data_L1,
               plot_period = "monthly", plot_export = FALSE)
```
<br>

The plot contains...

* the stem radius changes of the raw time-aligned `L1` dendrometer data in 
  the first panel, 
* the stem radius changes of the cleaned `L2` dendrometer data (with `L1` data 
  in the background in grey) in the second panel. Interpolated points are 
  circled and frost periods are indicated with a horizontal, cyan line
* the changes between `L1` and `L2` data (red) as well as the deleted values
  (pink) in the third panel
* and the tree water deficit (twd) in the last panel
<br>

When plotted monthly (`plot_period = "monthly"`), each change to the data 
gets an ID. The ID numbers facilitate the reversal of 
wrong or unwanted corrections. Alternatively, the plots can be drawn for 
the full period (`plot_period = "full"`) or for each year separately 
(`plot_period = "yearly"`). In these cases, the ID's are not displayed.
<br><br>

As a summary for each dendrometer series, an additional plot shows the yearly 
growth curves aligned by the day of year. The first year is shown in grey since
the measurements may start before or after the beginning of the growing season
and therefore the start of growth may not be comparable to other years.  
Beneath the plot all input values, some growth statistics as well as the 
package version used are reported.


```{r process to L2 last_plot, fig.asp = 1.4, fig.retina = 2, fig.keep = "last", echo = FALSE, results = "hide", message = FALSE}
proc_dendro_L2(dendro_L1 = dendro_L1, temp_L1 = temp_data_L1, 
               plot_period = "monthly", plot_export = FALSE)
```


## Adjustments to parameters of `proc_dendro_L2()`

### Rigidity of error detection
The rigidity of the error detection can be adjusted with the variables
`tol_out` and `tol_jump`. Both variables control the thresholds used for
the classification of outliers and jumps/shifts in the data.

If values are erroneously deleted (i.e. classified as outliers), the
value of `tol_out` should be increased (high values of `tol_out` decrease the 
rigidity of the thresholds). A decrease in the value of `tol_out`, in contrast, 
is appropriate if outliers are not detected as such (low values of `tol_out`
increase the rigidity of the thresholds).
Similarly, an increase in the value of `tol_jump` makes sense if a dendrometer
series is flattened during processing (see example below). A decrease, in
contrast, is appropriate if obvious jumps are not corrected.
<br>


#### Example of too rigid error detection
```{r shrinking example code, fig.asp=0.9, fig.show="hide", message=FALSE, warning=FALSE, results="hide"}
# Load dataset
data("dendro_data_L1")

# Subset dendrometer dataset for example
dendro_L1 <- dendro_data_L1 %>% 
  filter(series == "site-1_dendro-2")

# Clean dendrometer data (tol_jump = 1, tol_out = 10)
proc_dendro_L2(dendro_L1 = dendro_L1, temp_L1 = temp_data_L1, 
               tol_jump = 1, tol_out = 10, plot_export = FALSE)
```

```{r shrinking example plot, fig.asp = 0.5, fig.retina = 2, out.width = "70%", echo=FALSE}
data_shrink_L2 <- proc_dendro_L2(dendro_L1 = dendro_L1, 
                                 temp_L1 = temp_data_L1, 
                                 tol_jump = 1, tol_out = 10, plot = FALSE)

par(mar = c(2.1, 4, 1, 2))
plot(data = data_shrink_L2, value ~ ts, type = "n", 
     ylab = paste0("L2 (", "\u00b5", "m)"), las = 1)
lines(data = dendro_L1, value ~ ts, col = "grey70")
lines(data = data_shrink_L2, value ~ ts, col = "#08519c")
```
<br><br>

Such problems can be fixed by increasing the value of `tol_jump`.

<br>
```{r shrinking example higher tol code, fig.asp = 0.9, fig.show="hide", results="hide", message = FALSE, warning = FALSE}
# Clean dendrometer data (use default values of 'tol_jump' and 'tol_out')
proc_dendro_L2(dendro_L1 = dendro_L1, temp_L1 = temp_data_L1, 
               tol_jump = 50, tol_out = 10, plot_export = FALSE)
```

```{r shrinking example higher tol plot, fig.asp = 0.5, fig.retina=2, out.width = "70%", echo=FALSE}
data("dendro_data_L2")
dendro_L2 <- dendro_data_L2 %>% 
  filter(series == "site-1_dendro-2")

par(mar = c(2.1, 4, 1, 2))
plot(data = dendro_L2, value ~ ts, type = "n", 
     ylab = paste0("L2 (", "\u00b5", "m)"), las = 1)
lines(data = dendro_L1, value ~ ts, col = "grey70")
lines(data = dendro_L2, value ~ ts, col = "#08519c")
```
<br><br>


### Providing temperature data
Providing temperature data along with dendrometer data ensures that frost 
events are not treated as outliers during data cleaning. In periods of probable
frost (i.e. when the temperature < `lowtemp`) the threshold for outlier 
detection is multiplied by `frost_thr`.

Temperature data can either be provided along with the dendrometer data 
(series name has to  contain the string `temp`), or it can be provided 
separately to the argument `temp_L1` in the function `proc_dendro_L2()`.
<br><br>


#### Example without temperature data
If no temperature data is provided, a sample temperature dataset is used.
The sample temperature dataset assumes permanent frost in the months December,
January and February.

```{r frost no temp data_1, fig.asp = 0.9, fig.show="hide", results="hide", message=FALSE, warning=FALSE}
# Load dataset
data("dendro_data_L1")

# Subset dataset for example
dendro_L1 <- dendro_data_L1 %>% 
  filter(series == "site-1_dendro-4")

# Clean dendrometer data (without temperature data)
proc_dendro_L2(dendro_L1 = dendro_L1, plot_export = FALSE)
```

```{r frost no temp data_2, fig.asp = 0.5, fig.retina = 2, out.width = "70%", echo=FALSE, results="hide", warning=FALSE, message=FALSE}
# Process to L2 without temperature data
dendro_L2 <- proc_dendro_L2(dendro_L1 = dendro_L1, plot = FALSE)

par(mar = c(2.1, 4, 1, 2))
plot(data = dendro_L2, value ~ ts, type = "n", 
     ylab = paste0("L2 (", "\u00b5", "m)"), las = 1)
lines(data = dendro_L1, value ~ ts, col = "grey70")
lines(data = dendro_L2, value ~ ts, col = "#08519c")
```

The sample temperature dataset does not assume frost after the end of 
February, therefore many values during the frost shrinkage are classified as
outliers and deleted.
<br><br>


#### Example with temperature data
```{r frost with temp data_1, fig.asp = 0.9, fig.show="hide", results="hide", message=FALSE, warning=FALSE}
# Load time-aligned temperature data
data("temp_data_L1")

# Clean dendrometer data (with temperature data)
proc_dendro_L2(dendro_L1 = dendro_L1, temp_L1 = temp_data_L1, 
               plot_export = FALSE)
```

```{r frost with temp data_2, fig.asp = 0.5, fig.retina = 2, out.width = "70%", echo=FALSE, results="hide", warning=FALSE, message=FALSE}
dendro_L2 <- proc_dendro_L2(dendro_L1 = dendro_L1, temp_L1 = temp_data_L1, 
                          plot = FALSE)

par(mar = c(2.1, 4, 1, 2))
plot(data = dendro_L2, value ~ ts, type = "n", 
     ylab = paste0("L2 (", "\u00b5", "m)"), las = 1)
lines(data = dendro_L1, value ~ ts, col = "grey70")
lines(data = dendro_L2, value ~ ts, col = "#08519c")
# add abline to indicate frost
data_L2_frost <- dendro_L2 %>% 
  dplyr::rename(value_L2 = value)
plot_frost_period(df = data_L2_frost)
```

Including the temperature dataset increases the thresholds for jump and outlier
detection in periods of probable frost (i.e. when temperature < `lowtemp`), 
which are indicated with a horizontal cyan line in the plot.

The increase of the thresholds for outlier detection prevented the 
classification of the values during the frost shrinkage as outliers.
<br><br>


## Manual corrections
Remaining errors that cannot be removed by adjusting the default values
of `tol_jump` and/or `tol_out` can be corrected using the function
`corr_dendro_L2()`. This function can be used to reverse erroneous changes
or force changes that were not automatically made. There are three 
possibilities to manually correct remaining errors:

* **reverse**: specify the ID numbers of the changes that should be reversed.
  Remaining changes are renumbered starting at 1.
* **force**: force a shift in the data that was not corrected for by 
  specifying a date up to five days prior to where the shift should occur.
* **delete**: delete an entire period of erroneous data by specifying a
  date range. This can also be done for `L1` data with the function 
  `corr_dendro_L1()`.
<br>


#### Example of manual corrections
```{r corr_dendro_L2 example_1, fig.asp = 0.9, fig.keep = "first",  fig.retina=2, results="hide", message=FALSE, warning=FALSE}
# Load dataset
data("dendro_data_L1")

# Subset dataset for example
dendro_L1 <- dendro_data_L1 %>% 
  filter(series == "site-1_dendro-3")

# Clean dendrometer data
dendro_L2 <- proc_dendro_L2(dendro_L1 = dendro_L1, temp_L1 = temp_data_L1, 
                            plot_period = "monthly", plot_export = FALSE)
```
<br><br>

The following errors remained after data cleaning:

1. Corrections `59-61` are not necessary since the raw data seems to be correct
2. A jump occurring on August 13 is not corrected
3. Not all erroneous data is removed at the beginning (August 1-4)
<br>

Correction of the above-described errors:


```{r corr_dendro_L2 example_3, fig.asp = 0.9, fig.keep = "first", fig.retina=2, results="hide", message=FALSE, warning=FALSE}
# Correct described errors
corr_dendro_L2(dendro_L1 = dendro_L1, dendro_L2 = dendro_L2, 
               reverse = 59:61, 
               force = "2013-08-12", 
               delete = c("2013-08-01", "2013-08-04"),
               series = "site-1_dendro-3", plot_export = FALSE)
```

All three errors have been corrected successfully. The corrections are also
reflected in the returned `data.frame` and all changes are documented in the 
column `flags` as:

* `"rev"`: for reversed corrections with the argument `reverse`
* `"fjump"`: for forced jumps with the argument `force`
* `"del"`: for deleted values with the argument `delete`
<br><br>


# Step 3: Data aggregation (optional)
After error detection and processing and the manual removal of remaining 
errors, the package offers two functions to calculate additional physiological
parameters that may be of use for later analyses. The function `grow_seas` 
returns the day of year of growth onset and growth cessation. The function
`phase_stats` calculates the timing, duration, amplitude and the rate of change
of shrinkage and expansion phases.
<br>


### Growing season
The day of year of the start and end of the growing season can be calculated 
with the function `grow_seas`. 

```{r example grow_seas, results = "hide", message = FALSE, warning = FALSE}
# Load dataset
data("dendro_data_L2")

# Subset dataset for example
dendro_L2 <- dendro_data_L2 %>% 
  filter(series == "site-1_dendro-1")

# Calculate growing season start and end
grow_seas(dendro_L2 = dendro_L2)
```

The function returns a `data.frame` containing the day of year (`doy`) of the 
start and end of the growing season (Table 4). Values are 
returned starting from the second year only, since `gro_start` and `gro_end` 
depend on the values from the previous year.

```{r Table 4, echo = FALSE, message=FALSE, warning=FALSE}
# calculate grow_seas
grow_seas <- grow_seas(dendro_L2 = dendro_L2)
knitr::kable(grow_seas, caption = "**Table 4**: Sample output data of the 
             function `grow_seas`.")
```
<br><br>


### Phase statistics
Several characteristics of shrinkage and expansion phases can be calculated
with the function `phase_stats`. 

```{r example phase_stats, fig.show = "hide", results = "hide", message = FALSE, warning = FALSE}
# Load dataset
data("dendro_data_L2")

# Subset dataset for example
dendro_L2 <- dendro_data_L2 %>% 
  filter(series == "site-1_dendro-2")

# Calculate phase statistics
phase_stats(dendro_L2 = dendro_L2, plot_phase = TRUE, plot_export = FALSE)
```

The function returns a `data.frame` containing the timing, duration, amplitude
and slope of the shrinkage (`shrink`) and expansion (`exp`) phases (Table 5). 
In addition, the parameter `phase_class` identifies days on which radial change
is likely driven by transpiration (`phase_class = 1`) or temperature 
(`phase_class = -1`).

```{r Table 5, echo = FALSE, message=FALSE, warning=FALSE}
# calculate phase_stats
phase_stats <- phase_stats(dendro_L2 = dendro_L2)
knitr::kable(phase_stats[1:5, ], caption = "**Table 5**: Sample output data of 
             the function `phase_stats`.")
```
<br>

To evaluate the identification of shrinkage and expansion phases, all phases 
can be plotted by setting `plot_phase = TRUE`. The plot shows the 
maximum (filled circle) and minimum (filled triangle) of the respective phase
and reports its statistics. Empty circles and triangles show maxima or minima
of previous or subsequent phases.

```{r phase_stats figure, echo = FALSE, fig.keep = "first", fig.retina=2, results="hide", message=FALSE, warning=FALSE}
phase_stats(dendro_L2 = dendro_L2, plot_phase = TRUE, plot_export = FALSE)
```
<br>


The often reported parameter maximum daily shrinkage (`mds`) is equivalent to the
parameter `shrink_amp` in case `shrink_dur` is shorter than 24 hours. The 
following example demonstrates how `mds` can be extracted from the phase
statistics:
```{r extraction of mds, echo = TRUE, message=FALSE, warning=FALSE}
# Calculate phase statistics
phase_stats <- phase_stats(dendro_L2 = dendro_L2)

# Calculate maximum daily shrinkage (mds)
mds <- phase_stats %>% 
  mutate(mds = ifelse(shrink_dur < 1440, shrink_amp, NA))
```
<br><br>


# References
[Knüsel S, Peters RL, Haeni M, Wilhelm M, Zweifel R (2021) Processing and extraction of seasonal tree physiological parameters from stem radius time series. Forests: 12(6):765.](https://doi.org/10.3390/f12060765)

[Zweifel R, Haeni M, Buchmann N, Eugster W (2016) Are trees able to grow in 
periods of stem shrinkage? New Phytol. 211:839-49.](https://doi.org/10.1111/nph.13995)
<br>


# Citation
To cite `treenetproc` in a publication use:

```{r citation, echo = TRUE, warning=FALSE}
citation("treenetproc")
```
